Floating-Point Arithmetic on Deterministic Systems

### 2024-01-10 mss.tex
Update mss.tex

%% Configure first-page footer and later-page header.
\usepackage{fancyhdr}
\setlength{\headheight}{15.2pt}
\pagestyle{fancy}
%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              Manuscript submitted for review.\\
              ~ \\
              Address author correspondence to \texttt{davis68@illinois.edu, rshosted@illinois.edu}.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\it A Fixed and Unalterable Sound}}
\fancyhead[RO]{The Promise and Praxis of the Deseret Alphabet}
\fancyfoot[LE,RO]{\thepage}

@@

IEEE 754 specifies operations between numbers, including of different magnitudes.  The standard dictates behavior but not algorithm TODO
In practice, extended precision values are used in the intermediate steps of many algorithms, leading to greater accuracy than would otherwise be expected.

Since the IEEE 754 FP format packs values of different kind together bitwise, conventional integer operations such as \texttt{<<} left shift and \texttt{+} addition do not trivially apply.\footnote{For instance, left-shifting \texttt{1.0} does not result in \texttt{2.0} but \texttt{1.7014118e38}; we leave the mathematics of why as an exercise to the reader.}

@@

As an example, $16 + 4 \approx \texttt{add}(16, 4)$ is carried out as:
% https://cse.hkust.edu.hk/~cktang/cs180/notes/lec21.pdf
\begin{enumerate}
  \item  \begin{enumerate}
           \item  $16_{10} = \texttt{0100.0001.1000.0000.0000.0000.0000.0000}_{2}$
           \item  $ 4_{10} = \texttt{0100.0000.1000.0000.0000.0000.0000.0000}_{2}$
           \item  Larger exponent in $16_{10}$ = $\texttt{1000.0011}_{2}$ → $131 - 127 = 4$, $2^4$
           \item  $ \texttt{S}_{4_{10}} = \texttt{rsh}(\texttt{0100.0000.1000.0000.0000.0000.0000.0000}_{2}) = TODO here \texttt{0100.0000.1000.0000.0000.0000.0000.0000}_{2}$
         \end{enumerate}
  \item  \begin{enumerate}
           \item  Significand of $16_{10} = \texttt{[1]000.0000.0000.0000.0000}_{2}$
           \item  Significand of $ 4_{10} = \texttt{[1]000.0000.0000.0000.0000}_{2}$
           \item  $\texttt{S}_{16_{10}} + \texttt{S}_{4_{10}} = \texttt{[1]000.0000.0000.0000.0000}_{2} + \texttt{[1]000.0000.0000.0000.0000}_{2} = \texttt{1.0000.0000.0000.0000.0000}_{2}$
         \end{enumerate}
  \item  \begin{enumerate}
           \item  Significand:  $\texttt{S}_{\textrm{result}} = \texttt{rsh}(\texttt{1.0000.0000.0000.0000.0000}_{2}) = \texttt{1000.0000.0000.0000.0000}_{2}$
           \item  Exponent:  $\texttt{E}_{\textrm{result}} = \texttt{}_{2}\texttt{++}
           \item  Significand of $ 4_{10} = \texttt{[1]000.0000.0000.0000.0000}_{2}$
           \item  $\texttt{S}_{16_{10}} + \texttt{S}_{4_{10}} = \texttt{[1]000.0000.0000.0000.0000}_{2} + \texttt{[1]000.0000.0000.0000.0000}_{2} = \texttt{1.0000.0000.0000.0000.0000}_{2}$ $
         \end{enumerate}
\end{enumerate}

\begin{lstlisting}

\end{lstlisting}

BLAS rounding modes
https://bebop.cs.berkeley.edu/reproblas/

@@

\begin{enumerate}
  \item  The distance between two adjacent values changes based on the magnitude of the exponent and the distance from zero.
  \item  There is a relative approximation error for a given bitwidth in IEEE 754, called the {\emph machine epsilon}.\footnote{Note that this is different from the smallest representable value for a given bit width; e.g., for 32-bit single-precision \texttt{float} the smallest representable value is $\texttt{0000.0000.0000.0000.0000.0000.0000.0001} = 1\times 10^{-45}$.}
  \item  Operations between numbers of different magnitudes are particularly affected by their relative numerical horizon.
\end{enumerate}

\paragraph{Variable precision}.
For each value of the exponent \texttt{E}, the difference between two discrete values is determined by the relative magnitude of the significand \texttt{S}.  For instance, for values determined by the exponent $\texttt{E} = \texttt{011.1111.1}$, the difference between serial values is $\texttt{S} = \texttt{000.0000.0000.0000.0000.0001} \rightarrow 1.00000011920928955078125 - 1.0 = 0.00000011920928955078125 = 2^{-23}$.  For values determined by the exponent $\texttt{E} = \texttt{010.0000.0}$, the difference between serial values is TODO

% $2^{-23} \times 2^{-63} \approx 1.084 \times 10^{-19}$.
% SEEE.EEEE.EFFF.FFFF.FFFF.FFFF.FFFF.FFFF

This variable precision means that the precision of floating-point values varies across the range of representable numbers.  (See the paragraph on “operations beyond the horizon” below.)
% https://www.h-schmidt.net/FloatConverter/IEEE754.html

\paragraph{Machine epsilon}.
The machine epsilon for a particular bit width is determined by setting two to the negative power of the number of bits used for the magnitude of the mantissa; for 32-bit single-precision \texttt{float} this is $2^{-24} = \texttt{0000.0000.0000.0000.0000.0000.0000.0000}
$

\paragraph{Operations beyond the horizon}.
Operations between two numbers of fairly different precisions are particularly vulnerable to accuracy loss, although some numerical techniques can be employed to mitigate.  For example,

\begin{lstlisting}

\end{lstlisting}

iota, range, operations
nan

% Axelrod2006 https://www.jasss.org/9/4/4.html

Some numbers which do not 
0.1 etc. shift w/ bit loss

  Each 
iota, range, operations
nan

TODO

@@

As an example of the first, exception handling means that the hardware may specify rounding via an overflow flag TODO

As an example of the second, mixed-precision operations can yield unpredictable results depending on the compiler and hardware.  TODO

\begin{quote}
C support for signaling NaNs, or for auxiliary information that could be encoded in NaNs, is problematic.  Trap handling varies widely among implementations. Implementation mechanisms may trigger signaling NaNs, or fail to, in mysterious ways.  The IEC 60559 floating-point standard recommends that NaNs propagate; but it does not require this and not all implementations do.  \citep[p.~339]{Jones2008}
\end{quote}

The Vere runtime unifies NaNs.
quiet NaNs
specifically un/signed ambiguity

As an example of the third, fused multiply-add (FMA) ($a \times b + c$) is implemented on certain hardware to favor \texttt{double} operations and not quadruple-precision operations \citep[p.~5]{Kahan1997}.\footnote{While this particular example may not strike the reader as particularly egregious, the existence of deviations from the specification casts doubts on the reliability of any claimed adherence.}

rounding mode etc.
“Obtaining the correctly rounded result of an addition or subtraction operation requires an additional bit in the significand (as provided by the IEC 60559 guard bit) to hold the intermediate result” \citep[p.~65]{Jones2008}.
signed zero as nan
some have discussed the supercession of IEEE 754
five rounding modes even/away/zero/positive/negative

@@

EXAMPLE

“The Motorola DSP563CCC does not support NaNs or infinities.  Floating-point arithmetic operations do not overflow to infinity; they saturate at the maximum representable value” \citep[p.~338]{Jones2008}.

In a final buried admission, the ISO/IEC9899 C standard confesses its own fallibility:  “The floating-point model is intended to clarify the description of each floating-point characteristic and does not require the floating-point arithmetic of the implementation to be identical” \citep[fn.~21]{ISO/IEC9899:2018}.
% https://www.iso.org/obp/ui/#iso:std:iso-iec:9899:ed-4:v1:en

IEC 60559
subnormals, unnormals
“Some processors do not support subnormal numbers in hardware” \citep[p.~338]{Jones2008}.  (The risk is that this permits inadvertent division by zero.)  Various chipsets solve this exceptional behavior differently.

@@

\lstset{language=Hoon}
\begin{lstlisting}
::  call in code
(add:rs .1.2 .2.4)
\rightarrow
(add:ma .1.2 .2.4)
\rightarrow
add:%*(. ff w 8, p 23, b --127, r r)
\rightarrow
(bif (add:pa (sea a) (sea b)))

::  equivalent (unwrapped) operation in Hoon
=/  a=fn  []
TODO

::  equivalent Nock
TODO

::  result in binary (note that Urbit elides leading zeroes)
> `@ub`(add:rs .1.2 .2.4)  
0b100.0000.0110.0110.0110.0110.0110.0111

::  result in decimal
> `@rs`(add:rs .1.2 .2.4)  
.3.6000001
\end{lstlisting}

@@

\lstset{language=Hoon}
\begin{lstlisting}
++  add                                             ::  add; exact if e
  |=  [a=[e=@s a=@u] b=[e=@s a=@u] e=?]  ^-  fn
  =+  q=(dif:si e.a e.b)
  |-  ?.  (syn:si q)  $(b a, a b, q +(q))           ::  a has larger exp
  ?:  e
    [%f & e.b (^add (lsh [0 (abs:si q)] a.a) a.b)]
  =+  [ma=(met 0 a.a) mb=(met 0 a.b)]
  =+  ^=  w  %+  dif:si  e.a  %-  sun:si            ::  expanded exp of a
    ?:  (gth prc ma)  (^sub prc ma)  0
  =+  ^=  x  %+  sum:si  e.b  (sun:si mb)           ::  highest exp for b
  ?:  =((cmp:si w x) --1)                           ::  don't need to add
    ?-  r
      %z  (lug %fl a &)  %d  (lug %fl a &)
      %a  (lug %lg a &)  %u  (lug %lg a &)
      %n  (lug %na a &)
    ==
  (rou [e.b (^add (lsh [0 (abs:si q)] a.a) a.b)])
\end{lstlisting}
$ % TODO DELETEME after Obsidian

@@

For most purposes in the broader software world, tightly reproducible precision has not been a high priority.  Precision having already been sacrificed, the gist of the calculation is more important than the fourth decimal place (e.g. in realtime 3D graphics).

For example, consider the summation
on two different architectures with different optimization settings.

TODO
\lstset{language=C}
\begin{lstlisting}

\end{lstlisting}

@@

% TODO move this paragraph downchannel and mine Dawson
Others have recognized the challenge of the lack of determinacy before, such as the James Demmel and the \texttt{ReproBLAS} team \citep{Demmel2013, TODO} and \citeauthor{Dawson2013}.  (We discuss both in topical sections below.)
% https://bebop.cs.berkeley.edu/reproblas/docs/talks/SC13.pdf
% https://randomascii.wordpress.com/2013/07/16/floating-point-determinism/

@@

TODO
epoch system

@@

What about trimming floating-point values of their least significant bits?  When would this take place—at each step of a multi-step computation?  At the

There's a potentially computationally expensive step in checking and adjusting the FP value

@@

TODO
% https://www.ctrl-alt-test.fr/2018/making-floating-point-numbers-smaller/

@@

An optimized portable deterministic software library for floating-point calculations may be a sufficiently fast solution to meet Urbit's needs even for vector computations.  A different avenue worthy of investigation is to take IEEE 754 compliant floating-point values as inputs and outputs, then transform into a local representation for an optimized portable deterministic calculation.

@@

What degree of vetting will reliably answer the gap between IEEE 754 and hardware implementation for any particular operation?  \citet[pp.~330ff.]{Jones2008} and \citet{Goldberg2000} provide a careful analysis of accuracy errors inherent to IEEE 754 as a standard, but 
TODO only glancingly as a partial implementation.\footnote{See particularly the note on “Common Implementations” on p.~346 of \citet{Jones2008}.}

Having identified an appropriate subset of operations, we may imagine that the use of \texttt{#ifdef}, Autotools’ \texttt{configure}, and a jetting library may answer to our need.  Any jet library would have to be carefully constructed to avoid imposing tight discipline directly on the end user (modal Hoon author).  We cannot recommend this path today but do not consider the way to be shut.

@@

As an example of a refactoring of IEEE 754 operations for determinism, consider the \texttt{ReproBLAS} project (last update \asciitilde 2016.2.21).   \texttt{ReproBLAS} seeks to produce a set of reproducible deterministic algorithms reflecting the standard operations of BLAS.  It accomplishes this by introducing a binned data type and a set of basic operations carefully built on IEEE 754.
TODO
% https://bebop.cs.berkeley.edu/reproblas/index.php

@@

A fixed-point representation differs from a floating-point scheme in that the exponent is fixed by the protocol or metadata and thus only the sign and significand need be included in the bit representation.  (With an offset, even the sign can be elided.)  The advantage of such a scheme is that it affords the benefits of floating-point mathematics at near-integer operation speeds (e.g. left-shift to multiply by two).  One disadvantage is that there is a smallest representable value; this lack of subnormals requires either an underflow handler or the possibility of inadvertent division by zero.  Fixed-point operations could also be used as intermediates in calculations.  (This echoes once again the idea of conversion to an intermediate representation then conversion back out to IEEE 754.)

@@

@@

Lagoon\footnote{Linear AlGebra in hOON} is an Urbit library to facilitate Hoon-native mathematical operations.  It envisions six native types:

@@

\texttt{\%real} allows the rounding mode to be specified.  \texttt{\%fixp} requires an additional metadata statement of the expected precision.

@@

Lagoon has passed through several implementations.  The current reference implementation is the \texttt{lagoon} branch of the \texttt{https://github.com/urbit/urbit} repository \citep{Urbit}.

@@

\begin{enumerate}
  \item  Hardware FP on single machine for entire lifetime.
  \item  Optimized software FP.
  \item  Opaque calculation as callback.
  \item  Cached results by callsite.
  \item  Utilizing a subset of IEEE 754 in hardware.
  \item  Replacing IEEE 754 with another approach of sufficient speed, fixed-point and unum/posits chief among these.
\end{enumerate}

Several recent efforts on Urbit have encountered the difficulties of producing fast and reliable floating-point calculations on a Nock-based system.\footnote{UWasm, which as a WebAssembly implementation directly relies on floating-point computations\footnote{TODO.  See e.g. \citet{KloudKoder2022} for a discussion of the issues involved.}; \texttt{lazytrig}, which is slow and implemented purely in Hoon (Nock); Lagoon, which includes IEEE 754 operations with its linear algebra library; and Lora, TODO.}  We anticipate that, water finding its own level, each will adopt a suitable solution for the deterministic Nock platform.  We do not anticipate these to be the last foundational numerical libraries built on Urbit, but instead among the first.  Thus we have documented the paths we have explored as an annotated map for future travelers in search of a one true representation for continuous mathematics.
% KloudKoder2008 https://github.com/WebAssembly/design/issues/1456

>>>>

%% Configure first-page footer and later-page header.
\usepackage{fancyhdr}
\setlength{\headheight}{15.2pt}
\pagestyle{fancy}
%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              Manuscript submitted for review.\\
              ~ \\
              Address author correspondence to \texttt{~lagrev-nocfep}.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\it Urbit Systems Technical Journal}}
\fancyhead[RO]{Floating-Point Arithmetic on Deterministic Systems}
\fancyfoot[LE,RO]{\thepage}

@@

IEEE 754 specifies operations between numbers, including of different magnitudes.  The standard dictates behavior and provides outlines for arithmetic, but not leaves algorithmic details to the implementation.  Numbers are normalized by adjusting the exponent of the smaller operand and aligning the significands, then the operations are carried out.  In practice, extended precision values are used in the intermediate steps of many algorithms, leading to greater accuracy than would otherwise be expected.\footnote{“Each of the computational operations that return a numeric result specified by this standard shall be performed as if it first produced an intermediate result correct to infinite precision and with unbounded range, and then rounded that intermediate result, if necessary, to fit in the destination’s format.”  \citep{IEEE754-2008}  Note that, per \citeauthor{Risse2016}, “there is no indication whether or not a computation with IEEE 754 is exact even if all arguments are.”  The ISO/IEC9899 C standard confesses its own fallibility:  “The floating-point model is intended to clarify the description of each floating-point characteristic and does not require the floating-point arithmetic of the implementation to be identical” \citep[fn.~21]{IEC9899}.}
% https://www.iso.org/obp/ui/#iso:std:iso-iec:9899:ed-4:v1:en

Since the IEEE 754 floating-point format packs values of different kind together bitwise, conventional integer operations such as \texttt{<<} left shift and \texttt{+} addition do not trivially apply.\footnote{For instance, left-shifting \texttt{1.0} does not result in \texttt{2.0} but \texttt{1.7014118e38}; we leave the mathematics of why as an exercise to the reader.}

@@

@@

\begin{enumerate}
  \item  The distance between two adjacent values changes based on the magnitude of the exponent and the distance from zero.  (The significand resolution stays the same but the exponent changes.)
  \item  There is a relative approximation error for a given bitwidth in IEEE 754, called the {\emph machine epsilon}.\footnote{Note that this is different from the smallest representable value for a given bit width; e.g., for 32-bit single-precision \texttt{float} the smallest representable value is $\texttt{0000.0000.0000.0000.0000.0000.0000.0001} = 1\times 10^{-45}$.}
  \item  Operations between numbers of different magnitudes are particularly affected by their relative numerical horizon.
\end{enumerate}

\paragraph{Variable precision and truncation error}.

For most values of the exponent $\texttt{E}$, the difference between two discrete values is determined by the absolute magnitude of the significand $\texttt{S}$.  The difference between serial values is $\texttt{S} = \texttt{000.0000.0000.0000.0000.0001} \rightarrow 1.00000011920928955078125 - 1.0 = 0.00000011920928955078125 = 2^{-23}$.  This is multipled by the the result of the exponent $\texttt{E}$ and the bias, meaning that for each exponent value the difference between subsequent values changes.  (Figure \ref{fig:fp-res} represents this schematically.)

\begin{figure}[htb]
\includegraphics[width=\textwidth]{img/fp-res}
\caption{Schematic representation of granularity and variable precision of floating-point values and their relationship to the (continuous) set of real numbers.}
\label{fig:fp-res}
\end{figure}

However, for normalized numbers, or numbers that are left-shifted or right-shifted in order to carry out a calculation, values are determined by the {\emph relative shift} in exponent $\Delta\texttt{E}$.  For \texttt{E} = 2, for instance, the difference between serial values is $2^{-21}$.  This variable precision means that the precision of floating-point values varies across the range of representable numbers when operations take place.  Operations between two numbers of fairly different precisions are particularly vulnerable to accuracy loss, although some numerical techniques can be employed to mitigate.
% https://www.h-schmidt.net/FloatConverter/IEEE754.html

Truncation error results from terminating repeating “binaries” (by analogy with “decimals”).  Just as $\frac{1}{3} = 0.\bar{3} \approx 0.3333…$ has a finite precision when written in base-10, numbers that are not precise powers of two result in repeating fractions.  These necessarily terminate at the resolution of the significand.  The number and nature of truncation and rounding can significantly affect the accuracy of floating-point arithmetic and algorithms \citep{Izquierdo2006}.
% Izquierdo2006 https://www.jasss.org/9/4/4.html

\paragraph{Machine epsilon}.

The machine epsilon, or smallest value discernable from 1.0, is determined by the precision of the floating-point representation.  The machine epsilon for a particular bit width is determined by setting two to the negative power of the number of bits used for the magnitude of the mantissa and accounting for the leading implicit bit \texttt{1}; for 32-bit single-precision \texttt{float} this is $2^-{23-1} = 2^{-22}$.  Differences from 1.0 smaller than this cannot be represented in this bit width.

\paragraph{Sequence ordering}.

@@

\subparagraph{Optional aspects}

Several aspects of IEEE 754 are optional or advisory, including:

\begin{enumerate}
  \item  Exception handling means that the hardware may specify rounding via an overflow flag.
  \item  Extended precisions formats are not a huge deal to leave out, but extended precision arithmetic (used for intermediate results) can materially change results.
  \item  Subnormals are optional; some platforms may flush them to zero or (worse) allow subnormal support to be disabled in certain cases.\footnote{“Some processors do not support subnormal numbers in hardware” \citep[p.~338]{Jones2008}.  (The risk is that this permits inadvertent division by zero.)  Various chipsets solve this exceptional behavior differently.}
\end{enumerate}

\subparagraph{Omissions}

Whether something is a gap or optional is something of a philosophical question for us, but some parts are underspecified in a way that makes portability impossible.  E.g., mixed-precision operations can yield unpredictable results depending on the compiler and hardware.  This is a function of rounding modes, precision loss, precision of intermediate results, and the presence or absence of dedicated hardware support for certain precision combinations.

\subparagraph{Inexact implementation}

Failure to implement IEEE 754 would most likely happen inadvertently (as with the Pentium FDIV bug in the 1990s).  Chipset designers may also deviate from this behavior.

For instance, IEEE 754 defines a range of numbers as “not-a-number” values, or NaNs.  Per the specification, a NaN can be a signalling NaN, meaning that it intends to flag and possibly disrupt a problematic computation;\footnote{“C support for signaling NaNs, or for auxiliary information that could be encoded in NaNs, is problematic.  Trap handling varies widely among implementations. Implementation mechanisms may trigger signaling NaNs, or fail to, in mysterious ways.  The IEC 60559 floating-point standard recommends that NaNs propagate; but it does not require this and not all implementations do.”  \citep[p.~339]{Jones2008}} or a quiet NaN, which does not raise such an exception and merely yields a result with the NaN propagated to the final result.\footnote{In Urbit, the Vere runtime unifies NaNs, meaning that any bitwise information which may be encoded in the significand field—the “NaN payload”—is thrown away.}  Not all processors implement this part of IEEE 754 correctly:  “The Motorola DSP563CCC does not support NaNs or infinities.  Floating-point arithmetic operations do not overflow to infinity; they saturate at the maximum representable value” \citep[p.~338]{Jones2008}.

As a further example, fused multiply-add (FMA) ($a \times b + c$) is implemented on certain hardware to favor \texttt{double} operations and not quadruple-precision operations \citep[p.~5]{Kahan1997}.\footnote{While this particular example may not strike the reader as particularly egregious, the existence of deviations from the specification raises the specter of nondeterminism.}

\subparagraph{Out-of-sequence computations}

@@

As demonstrated above, out-of-sequence or resequenced computations can affect results due to rounding behavior and the “numerical horizon” which results between values.  These can happen due to multithreaded computation or an optimizing compiler.

\paragraph{Rounding mode}.

IEEE 754 floating-point operations take place using one of four rounding modes.

\begin{enumerate}
  \item  Round to nearest, ties to even.  Set ties to the last bit as zero (even).  The default.
  \item  Round to zero.  Truncate, effectively rounding positive numbers down and negative numbers up.
  \item  Round toward positive infinity.  Up regardless of sign.
  \item  Round toward negative infinity.  Down regardless of sign.
\end{enumerate}

The rounding mode can affect the result of computations, and if other processes are changing the mode (which can even be set per-thread), results may not be reliably reproducible.

“Obtaining the correctly rounded result of an addition or subtraction operation requires an additional bit in the significand (as provided by the IEC 60559 guard bit) to hold the intermediate result” \citep[p.~65]{Jones2008}.

@@

@@

\lstset{language=Hoon}
\begin{lstlisting}
++  add                                             ::  add; exact if e
  |=  [a=[e=@s a=@u] b=[e=@s a=@u] e=?]  ^-  fn
  =+  q=(dif:si e.a e.b)
  |-  ?.  (syn:si q)  $(b a, a b, q +(q))           ::  a has larger exp
  ?:  e
    [%f & e.b (^add (lsh [0 (abs:si q)] a.a) a.b)]
  =+  [ma=(met 0 a.a) mb=(met 0 a.b)]
  =+  ^=  w  %+  dif:si  e.a  %-  sun:si            ::  expanded exp of a
    ?:  (gth prc ma)  (^sub prc ma)  0
  =+  ^=  x  %+  sum:si  e.b  (sun:si mb)           ::  highest exp for b
  ?:  =((cmp:si w x) --1)                           ::  don't need to add
    ?-  r
      %z  (lug %fl a &)  %d  (lug %fl a &)
      %a  (lug %lg a &)  %u  (lug %lg a &)
      %n  (lug %na a &)
    ==
  (rou [e.b (^add (lsh [0 (abs:si q)] a.a) a.b)])
\end{lstlisting}

@@

For most purposes in the broader software world, tightly reproducible precision has not been a high priority.  Precision having already been sacrificed, the gist of the calculation is more important than the fourth decimal place (e.g. in realtime 3D graphics).  This leads the phrase “implements the IEEE 754 standard” to be interpreted erroneously to imply full reproducibility \citep{Figueroa2000}.
% S. A. Figueroa del Cid. A Rigorous Framework for Fully Supporting the IEEE Standard for Floating-Point Arithmetic in High-Level Programming Languages. PhD thesis, New York University, Jan. 2000.

For example, consider the expression $(a \times b) + c$.  If a compiler permits the two operations to be evaluated sequentially (a multiplication followed by an addition), then rounding occurs twice.  If a compiler optimizes the operation into an FMA, or fused multiply-add, then a single rounding occurs.  \citeauthor{Peters2021} presents a pathological case for 32-bit single-precision floating-point values:  $a = 1.00000011920929$, $b = 53400708$, and $c = -b$.  In this case, the two-stage operation wipes out the $0.00000011920929$ component of $a$, yielding $a$ as an integer.  Then $c$ is added and the result is $8$.  With FMA as a single-step operation, the (correct) answer $6.365860462$ is obtained.  The optimization is more correct than the naïve route in this case.
% https://momentsingraphics.de/FMA.html

However, in another example due to \citeauthor{Dawson2013}, FMA yields incorrect results:  for $a \times b + c \times d$ with $a = c$ and $b = -d$, the answer should be zero, and calculated in two steps will typically be zero.  With a fused multiply-add, however, the code becomes \texttt{fmadd(a, b, c*d)}, rounding the multiplication of $c$ and $d$ but not that of $a$ and $b$; the answer will likely not be zero.

The situation grows more ambiguous across architectures.  \citet[p.~346]{Jones2008} presents the pathological case of a compliant platform that may use extended precision bits in the calculation of \texttt{a + b}:

\lstset{language=C}
\begin{lstlisting}
#include <stdio.h>

extern double a, b;

void f(void) {
    double x;
    x = a + b;
    if (x != a + b)
        printf("x != a + b\n");
}
\end{lstlisting}

In this hypothetical case,“any extended precision bits will be lost in the first calculation of a+b when it is assigned to x. The result of the second calculation of a+b may be held in a working register at the extended precision and potentially contain additional value bits not held in x, the result of the equality test then being false.”

\citet{Higham2002} provides further examples of pathological cases.
%  HIGHAM N J (2002) _Accuracy and Stability of Numerical Algorithms_, (second ed.), Philadelphia, USA: SIAM.

@@

The challenge of the lack of determinacy for certain critical applications has been acknowledged before, such as by James Demmel and the \texttt{ReproBLAS} team \citep{Demmel2013, ReproBLAS} and by \citeauthor{Dawson2013}.  Dawson makes much of the effect of rounding modes and the option to disable subnormals, both of which would have major effects on computational reproducibility.  The situation is in fact worse for transcendental functions, because there is necessarily truncation and/or rounding error
% https://bebop.cs.berkeley.edu/reproblas/docs/talks/SC13.pdf
% https://randomascii.wordpress.com/2013/07/16/floating-point-determinism/

@@

The Urbit runtime provides an epoch system, meaning that the event log is separated into snapshots and subsequent events \citep{UrbitEpoch}.  This is currently used to monitor the use of old binaries which could potentially have a jet mismatch.  It would be moderately straightforward to extend this functionality to record the compilation flags and architecture of that Vere binary, which could be useful in event playback.  However, this remains an unsatisfactory solution because it would lead to Urbit runtime instances intentionally producing different code (rather than a jet mismatch which would require correction).
% https://roadmap.urbit.org/project/epoch-system

@@

What about trimming floating-point values of their least significant bits?  When would this take place—at each step of a multi-step computation?  At the level of single-bit rounding errors, this would potentially work, and amounts to selecting a rounding mode towards even (last digit \texttt{0}).  Accrual across multiple calculations could potentially render this unreliable, particularly if different computational paths are supposed to lead to the same result and do not as a result of docking.

One could also envision docking more than the last bit.  This introduces a step to check and adjust the floating-point value, and in addition breaks IEEE 754 compliance—at which point the trouble of trying to reconcile IEEE 754 with determinism fails.

@@

A related technique could pack bits of larger floating-point values into smaller ones, but this is functionally a software-defined solution (see, e.g., \citet{LeBrun2018}).
% https://www.ctrl-alt-test.fr/2018/making-floating-point-numbers-smaller/

@@

An optimized portable deterministic software library for floating-point calculations may be a sufficiently fast solution to meet Urbit's needs even for vector computations.  A different avenue worthy of investigation is to take IEEE 754 compliant floating-point values as inputs and outputs, then transform into a local representation for an optimized portable deterministic calculation.  For instance, \citet{Thall2007} presents the concept of “unevaluated sums”, a generalized technique for accruing error in situations where additional precision is necessary for accuracy.  However, even with an agreed-upon standard library like SoftFloat, it is important to keep in mind that results like transcendental functions are still not correctly known.\footnote{To correctly calculate a trigonometric function for \texttt{double} may take over a hundred bits of precision before correct rounding can be determined.  Furthermore, the C \texttt{math.h} implementation of $sin$ may or may not use \texttt{fsin}.}  This particular poses a problem for functions like $sin$ which may be calculated by different routes in Hoon/Nock and in C/Rust.  For the time being, we conclude that Urbit's discipline requires only using Hoon/Nock implementations of transcendental functions.
% Thall, Andrew (2007). Extended-Precision Floating-Point Numbers for GPU Computation. Technical Report CIM-007-01. The University of North Carolina at Chapel Hill. [Author's version](http://andrewthall.org/papers/df64_qf128.pdf)

@@

What degree of vetting will reliably answer the gap between IEEE 754 and hardware implementation for any particular operation?  \citet[pp.~330ff.]{Jones2008} and \citet{Goldberg1991} provide a careful analysis of accuracy errors inherent to IEEE 754 as a standard, but due to the variety of possible scenarios do not treat of real compilers and chipsets much.\footnote{See particularly the note on “Common Implementations” on p.~346 of \citet{Jones2008}.}  Trivially, as demonstrated above in the Python example, $(a + b) + c \neq a + (b + c)$, and even modest reordering of operations by a zealous compiler optimization is susceptible of introducing nonportable and thus nondeterministic (in our sense) behavior.

Having identified an appropriate subset of operations, we may imagine that the use of \texttt{#ifdef}, Autotools’ \texttt{configure}, and a jetting library may answer to our need.  Any jet library would have to be carefully constructed to avoid imposing tight discipline directly on the end user (modal Hoon author).  We cannot recommend this path today but do not consider the way to be shut, especially given liberal use of \texttt{volatile}.

@@

As an example of a refactoring of IEEE 754 operations for determinism, consider the \texttt{ReproBLAS} project (last update \asciitilde 2016.2.21).   \texttt{ReproBLAS} seeks to produce a set of reproducible deterministic algorithms reflecting the standard operations of BLAS \citep{ReproBLAS}.  It accomplishes this by introducing a binned data type and a set of basic operations carefully built on IEEE 754 for the objective of completely portable reproducibility.\footnote{“Using our default settings, reproducibly summing n floating point types requires approximately $9n$ floating point operations (arithmetic, comparison, and absolute value). In theory, this count can be reduced to 5n using the new "augmented addition" and "maximum magnitude" instructions in the proposed IEEE Floating Point Standard 754-2018.”  \citep{ReproBLAS}}  This is similar to our proposal for a vetted jetting library and may be worth attention, particularly in association with requirements around \texttt{-O0}.
% ReproBLAS https://bebop.cs.berkeley.edu/reproblas/index.php

@@

A fixed-point representation differs from a floating-point scheme in that the exponent is fixed by the protocol or metadata and thus only the sign and significand need be included in the bit representation.  (With an offset, even the sign can be elided.)  The advantage of such a scheme is that it affords the benefits of floating-point mathematics at near-integer operation speeds (e.g. left-shift to multiply by two).  One disadvantage is that there is a smallest representable value; this lack of subnormals requires either an underflow handler or the possibility of inadvertent division by zero.  Fixed-point operations could also be used as intermediates in calculations.  (This echoes once again the idea of conversion to an intermediate representation then conversion back out to IEEE 754.)\footnote{“One solution to implementing floating-point types on processors that support fixed-point types is to
convert the source containing floating-point data operations to make calls to a fixed-point library” \citep[p.~346]{Jones2008}.  Note that the sense of our current interest is reversed.}

@@

\subsection{Irregularities}

Any approach to modeling real numbers runs the risk that different calculation pathways will yield a different kind of inexactness in the result.  These can be mitigated by some of the approaches suggested above, and also by checking the correspondence of the Hoon code and the underlying jet, particular for known edge cases in behavior.  While Hoon–jet compliance is an open research problem,\footnote{Particularly as regards co-generation of Hoon and C/Rust, or formal proofs.} we can apply principles of unit testing together with a period of testing Nock and jet compliance.\footnote{The Vere runtime supports a debugging flag which runs both the Nock and the jet and checks for identical results.}

Jet mismatches have been rare in the current era.\footnote{By “current era”, we mean after the last global network breach in \asciitilde 2020.12 \citep{UrbitBreach}.}  Some jet “mismatches” occur because the runtime raises a different error than the corresponding Hoon—these are relatively innocuous.  Others may occur because actually different results are produced for different input.  These are grave, and ultimately motivated the introduction of the epoch system so that event log replays can take into account the previous less-perfect jet version in the runtime \citep{UrbitEpoch}.
% UrbitBreach https://roadmap.urbit.org/project/last-network-breach

@@

Lagoon\footnote{Linear AlGebra in hOON} is an Urbit library to facilitate Hoon-native mathematical operations.  It envisions six native types,

@@

for which \texttt{\%real} allows the rounding mode to be specified; and \texttt{\%fixp} requires an additional metadata statement of the expected precision.

@@

Lagoon has passed through several implementations and remains in active development.  The current implementation is the \texttt{lagoon} branch of the \texttt{https://github.com/urbit/urbit} repository \citep{Urbit}.

@@

\begin{enumerate}
  \item  Hardware FP on single machine for entire lifetime.
  \item  Optimized software FP with vetted jetting library.
  \item  Opaque calculation as callback.
  \item  Cached results by callsite.
  \item  Utilizing a subset of IEEE 754 in hardware.
  \item  Replacing IEEE 754 with another approach of sufficient speed, fixed-point and unum/posits chief among these.
\end{enumerate}

Several recent efforts on Urbit have encountered the difficulties of producing fast and reliable floating-point calculations on a Nock-based system.\footnote{UWasm, which as a WebAssembly implementation directly relies on floating-point computations \citep{UWasm} (see e.g. \citet{KloudKoder2022} for a discussion of the issues involved); \texttt{lazytrig}, which is slow and implemented purely in Hoon (Nock); Lagoon, which includes IEEE 754 operations with its linear algebra library ({\emph q.v.}); and Lora, TODO.}  We anticipate that, water finding its own level, each will adopt a suitable solution for the deterministic Nock platform.  We do not anticipate these to be the last foundational numerical libraries built on Urbit, but instead among the first.  Thus we have documented the paths we have explored as an annotated map for future travelers in search of a one true representation for continuous mathematics.
% UWasm https://github.com/Quodss/urwasm
% KloudKoder2008 https://github.com/WebAssembly/design/issues/1456
